---
title: "Theophylline dataset"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Theophylline dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
library(DiagrammeR)
```

## Theophylline

The data in Theoph.csv came from a study of the kinetics of the anti-asthmatic
agent theophylline reported by Boeckmann et al. (1992). In this experiment,
the drug was administered orally to twelve subjects, and serum
concentrations were measured at ten time points per subject over the subsequent
25 hours.

The common model for the kinetics of theophylline following oral administration is
a one-compartment model with first-order absorption and elimination. In this
model, the subject-specific pharmacokinetic parameters to be estimated are -

* the absorption rate constant, $ka$,
* the volume of distribution, $V$, and
* the elimination rate constant, $k$.

```{r echo=FALSE}
mermaid("
        graph LR
Depot --ka --> Central((Central))
Central((Central)) -- k --> Out
        ")
```

Notes -

1. $Dose$ (units - milligram) is given at time $t = 0$ into the $Depot$ compartment - represents the gut
2. The $Dose$ moves from the $Depot$ compartment to the $Central$ compartment with a first-order absorption rate constant $ka$
3. $Central$ compartment is the observation compartment where concentrations of the drug $C_p(t)$ are measured (units - milligram/Liter).
4. Volume of distribution parameter $V$ (units = Liter) is a proportionality constant that scales the amount of drug in the $Central$ compartment (mg) to the observed concentration (mg/L) - $Concentration = Dose/V$
5. $k$ (units - 1/time) is the parameter that describes the elimination rate constant of the drug from the $Central$ compartment
6. $CL$ (units - Liter/time) is useful parameter that relates the rate of elimination (units - milligram/time) to the measured concentration.
$$ Rate of Elimination = CL \times C_p $$

$CL$ can also be derived from $k$, $V$ where,
$$ CL = k \times V$$

7. $Out$ is collection compartment without any exit rates. Such a compartment can be use to accumulate drug over time and is commonly referred to as area under the curve or $AUC$.  
$$ AUC  = \int\limits_0^\infty C_p\, \mathrm{d}t$$

#### Analytical Solution

$$C_p(t) = \frac{ Dose }{ V }\frac{ ka }{ ka - k }(e^{-k(t-t_D)} - e^{-ka(t-t_D)}) $$

#### ODE

$$
\begin{array}{lcl} \frac{ dDepot }{ dt } & = & -ka\times Depot \\ \frac{ dCentral }{ dt } & = & ka\times Depot - k\times t \end{array}
$$

## Data Structure

1. Subject
an ordered factor with levels 1, ..., 12 identifying the subject on whom the observation was made. The ordering is by increasing maximum concentration of theophylline observed.

2. Wt
weight of the subject (kg).

3. Dose
dose of theophylline administered orally to the subject (mg/kg).

4. Time
time since drug administration when the sample was drawn (hr).

5. conc
theophylline concentration in the sample (mg/L).